---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BLA

<!-- badges: start -->
<!-- badges: end -->

## Overview

The BLA `R` package provides a series of tools to fit boundary line models to a 
data set as proposed by Webb (1972). It includes a suite of methods which have
been introduced since the original manually drawn boundary lines were
proposed. These include methods based on binning the independent
variable, the BOLIDES algorithm of Schug et al. (1995), quantile regression
and the censored bivariate normal model of Milne et al. (2006). It also houses data exploration techniques to check for outliers and to provide initial evidence for a limiting boundary in data sets as an initial steps before doing boundary line analysis. It also includes functions to determine suitable starting guess values for use in fitting various boundary line models most of which use numerical optimization procedures. Learn more in `vignette("Introduction to boundary line analysis")`.

## How can I installation the package?

Install released version from CRAN:

``` {r, include = TRUE, eval=FALSE}
install.packages("BLA")
```

## What can I use the BLA package for?

If you have data on pH and crop yield collected over several farms in a non experimental survey (other factors not controlled), you can determine the largest expected yields for a given pH value, also called the boundary pH value. various 
methods to fit the boundary model including the `blbin()`, `BOLIDES()`, `blqr()` and `cbvn()`. The example below uses the `BOLIDES()` function:

```{r}
library(BLA)
library(aplpack)

# 1. Outlier detection using bagplot() function
nobs<-length(soil$pH)
vals_ur<-matrix(NA,nobs,2)
vals_ur[,1]<-soil$pH 
vals_ur[,2]<-soil$yield

bag<-bagplot(vals_ur,create.plot = F )
vals<-rbind(bag$pxy.bag,bag$pxy.outer) # new excludes bivariate outliers

# 2. Fit the boundary model in form of a linear plateau ("lp")

x<-vals[,1]
y<-vals[,2]
theta =c(-12.99,2,13 ) # initial start values for optimisation

model<-BOLIDES(x,y, theta = theta,model = "lp",xlab=expression("pH"), 
        ylab=expression("Yield/ t ha"^-1), pch=16,
       col="grey40", cex=0.8)

model

```

The boundary yield given the pH for each farm can be predicted:

```{r}
pH_values<-soil$pH
pH_values[which(is.na(x)==T)]<-mean(x,na.rm=T)
predicted_yield<-predictBL(model,pH_values)

head(predicted_yield) # predicted yield for the first six fams
```

The critical pH value using the model parameters can be determined.
This is the pH beyond which yield increase response is not expected.

```{r}
intercept<-model$Parameters[1]
slope<-model$Parameters[2]
plateau<-model$Parameters[3]

critical_pH<- (plateau-intercept)/slope

print(critical_pH)
```

Other boundary line post-hoc analysis procedures can be conducted. For more information, see the `vignette("Introduction to boundary line analysis")` and `vignette("Use of censored bivariate normal model function for yield gap analysis")`
