---
title: "Use of censored bivariate normal model function for yield gap analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use of censored bivariate normal model function for yield gap analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The censored bivariate normal model (Lark and Milne, 2016) is a statistical method of fitting the boundary line to data. It fits the boundary line using the maximum likelihood approach on a censored bivariate distribution. This removes the subjectivity mostly associated with the other boundary line fitting methods ( BOLIDES, Binning and quantile regression) of selecting boundary points to which the boundary line is fitted. It also gives evidence for presence of boundary in a data set. This method has been previously used to fit boundary lines to data on nitrous oxide emission as a function of soil moisture (Lark and Milne, 2016) and, Wheat yield as a function of Nutrient concentration (Lark *et al.* 2020). An `R` package, `BLA`, contains exploratory and boundary fitting functions that can be used in the process of fitting a boundary line to data using censored bivariate normal model. This vignette gives a step by step process for fitting the boundary line and subsequent post-hoc analysis. In this demonstration, the data set soil will be used and boundary line analysis will be carried out on soil P and pH.

# Installing dependent packages

The `BLA` package requires that the packages `MASS,` `aplpack` and `mvtnorm` are installed in `R`. You can install these packages using the following code

```{r, eval=FALSE}
# package names
packages <- c("MASS", "mvtnorm", "aplpack")
#
# install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)){
   install.packages(packages[!installed_packages])}

```

The `BLA` package can then be loaded

```{r setup}
library(BLA)
library(aplpack)
```

# Yield gap analysis using boundary lines

The data that is going to be used for this is called the `soil`which is part of the `BLA` package. It contains measures of wheat yield on farms from different parts of the UK and its associated soil properties pH and phosphorus. The process of yield gap analysis involves two step that include (1) fitting the boundary line and (2) determining the most limiting factor. We will first fit the boundary line to soil P and then to pH after which the most limiting factor will be determined.

```{r}
data("soil")
```

## Soil P boundary line fitting

### Data exploration

Exploratory analysis is an essential initial step in fitting a boundary line to data. This step ensures that the assumptions for the censored bivariate normal model are met. Three exploratory procedures are performed on the data which include (1) Testing for normality of variables, (2) Removal of outliers, and (3) Testing for evidence of boundary in the data.

1\. Testing for normality of independent and dependent variable

Boundary line model fitting using `cbvn()` requires that the independent (`x`) and dependent (`y`) variables are normally distributed. The `summa()` function gives indices of the distribution of a variable. A variable is assumed to be normally distributed if the skewness index is between -1 and 1. The `summaplot()` function gives a visual description of the distribution.

```{r,fig.align='center', fig.dim=c(4,4)}
x<-soil$P
y<-soil$yield

#Distribution of the x variable

summa(x) # Gives indices for skewness. Normal distribution range of -1 to 1. If not normally distributed,a transformation is required 
summaplot(x) # histogram and qqplot for distribution of x

summa(log(x)) 
summaplot(log(x)) 

#Distribution of the y variable

summa(y)# Gives indices for skewness. Normal distribution range of -1 to 1. If not normally distributed,a transformation is required 
summaplot(y) # histogram and qqplot for distribution of y
```

2\. Removal of outliers

Boundary line analysis is sensitive to outlying values and hence it is required that they are identified and removed. A bag plot (a bivariate box plot) is used to identify outliers. The bag plot has four main components, (1) a depth median (equivalent to the median in a box plot) which is represents the centre of the data set, (2) The bag which contains 50% of the data points (eqiuavalent to the interquartile range), (3) a 'fence' that separates probable outliers, (4) a loop which contains points outside the bag but are not outliers.

```{r,fig.align='center', fig.dim=c(4,4)}
x<-log(x) # since we requred a transformation
df<-data.frame(x,y) 
bag<-bagplot(df,show.whiskers = FALSE, ylim=c(0,20))
legend("topright", c("bag","loop","outliers", "d.median"), 
           pch = c(15,15,16,8),col=c("blue","lightblue","red","red"),cex=0.7)

# Combine data points from "bag" and within the loop

dat<-rbind(bag$pxy.bag,bag$pxy.outer)

# Output is a matrix, we can pull out x and y variables for next stage 

x<-dat[,1] 
y<-dat[,2]
```

3\. Testing data for presence of boundary

It is important to note that not all data are suitable for boundary line analysis. Data used in boundary line analysis should exhibit some form of boundary on its edges (to satisfy the assumption that this is response when other factors are not limiting). If a boundary exists in a data set, it is expected that points at the upper edges of the data cloud will be close to each other compared to a bivariate normally distributed data for which data points at the upper edges are only due to random error. A boundary can be assumed in a data set if there is evidence of clustering at the upper edges. The `expl_boundary()` function, which is based on the convex hull, can be used to access presence of boundary (Milne et al, 2006). This function checks probability of the observed clustering if it came from a bivariate normal distribution (p-value).

```{r, eval=FALSE,fig.align='center', fig.dim=c(4,4)}

expl_boundary(x,y) # may take a few minutes to complete

```

From the results, the probability (p-value) of the having points close to each other as in our data if it came from a bivariate normally distributed data is less than 5%. Therefore, there is evidence of bounding effects in the data in the right and left sections of the data. Note that, in the plot, the data is split into right and left sections to get more information on the clustering nature of points.

### Fitting the boundary line to data

From the exploratory tests, they indicated that the data provides evidence of boundary existence, outliers have been identified and removed, and the variables *`x`* and *`y`* are normally distributed. We therefore, proceed to fit a boundary line model to the data set using the censored bivariate normal model. The function *`cbvn()`* fits the boundary line to the data. For more information about the arguments of this function, check

```{r, eval=FALSE}
?cbvn
```

Argument values for the function *`cbvn()`* need to be set . Firstly, create a data-frame containing *x* and *y,* called `vals`.

```{r}
vals<-data.frame(x,y) #This is an input dataframe containing the variables
```

Secondly, the *`cbvn()`* requires initial starting values, *`theta`*, which are parameters of the boundary line (the censor) and the bivariate normal distribution. Starting values of the boundary line depend on the model that one wishes to fit to the data (see options in *`?cbvn`*) . In this case, we shall fit a linear+Plateau model (lp) and hence the parameters are the plateau value and the intercept and slope of the linear component. The boundary line start values can be obtained using the function *`startValues()`.* For more information on the *`startValues()`* function run the code *`?startValues`*. With a scatter plot of *y* against *x* active in the plot window in `R`, run the function *`start.values("lp")`*, then click on the plot, the point you expect to be the lowest and the largest response of a linear model at the boundary. Note that starting values for other models can be determined by changing the model name from "lp" to the desired models (see *`?startValues`* help file).

```{r, eval=FALSE}
plot(x,y)

startValues("lp") 
```

Next, determine the parameters of the bivariate normal distribution which include the means of the *x* and *y* variables, standard deviation of the *x* and *y* variables, and the correlation of *x* and *y*.

```{r}
mean(x) 
mean(y) 
sd(x) 
sd(y) 
cor(x,y)

#The parameters of the boundary line and the data can be combined in a #vector theta in the order 
#theta<-c(intercept, slope, max response, mean(x), mean(y), sd(x), sd(y), cor(x,y))

theta<-c(4,3,13.6,3,9,0.5,1.9,0.05) 
```

Another important argument is the standard deviation of the measurement error, *`sigh`*. This value can be obtained from standard deviation of repeated measurements of the sample if this data is available. However, in cases when it is not available, it can be estimated using different options. One option of estimation is to use a variogram if the location data (*x y* co-ordinate) is available. In that case, nugget variance which is the unexplained short distance variations can be taken as an estimate of the measurement error. If this is not possible, a profiling procedure can be done. This is done by fitting a proposed model using a varied number of *`sign`* values while keeping the rest of the model parameters constant. The log-likelihood values of the model for each *sign* are determined and the value that maximizes the likelihood can be selected from the log-likelihood profile. This can be implemented using the *`ble_profile()`* function. For more information on this function run

```{r, eval=FALSE}
?ble_profile
```

The possible `sigh` values can be set to 0.5, 0.7 and 0.8 and we can check the likelihood profile.

```{r, eval=FALSE,fig.align='center', fig.dim=c(4,4)}

sigh=c(0.5,0.7,0.8)
ble_profile(vals,theta,sigh,model = "lp") # may take a few minutes to run for large datasets
```

From the likelihood profile, the `sigh` that maximizes the likelihood is around 0.65. We can use this for the value of *`sigh`*. Now that all the arguments required for `cbvn()` function are set, the boundary line can be fitted

```{r, eval=TRUE,fig.align='center', fig.dim=c(4,4) }

model1<-cbvn(vals,theta,sigh=0.7,model = "lp", xlab=expression("P /log ppm"), ylab=expression("Yield /t ha"^{-1}), pch=16, col="grey")

model1
```

The boundary line is fitted and the parameters together with their corresponding standard error values are obtained. However, there is one question that needs to be addressed before we proceed,

Is the boundary line ideal for this data?

While fitting the BL model to the data, the *`cbvn()`* also fits a bivariate normal model with no boundary and calculates it `AIC` value. From our output, the `AIC` value of the BL model is lower than that of the bivariate normal model. Therefore, the BL model is appropriate. The parameters of the boundary line can now be used to predict the boundary yield given the value of P.

The function `predictBL()` in the BLA package can be used for this. For more information about the function, see

```{r, eval=FALSE}
?predictBL
```

We need to predict the largest expected yield for each point given the soil P. We will replace any missing values with the mean value of P.

```{r, eval=TRUE}

xp<-log(soil$P) # let xp be the P content in our dataset
xp[which(is.na(xp)==T)]<-mean(xp,na.rm=T)
P<-predictBL(model1,xp)
```

We can proceed to the next variable pH using the same procedure.

## Soil pH boundary line fitting

### Data exploration

1\. Testing for normality of independent and dependent variable

```{r,fig.align='center', fig.dim=c(4,4)}
x<-soil$pH
y<-soil$yield

#Distribution of the x variable

summa(x) # Gives indices for skewness. Normal distribution range of -1 to 1. If not normally distributed,a transformation is required 
summaplot(x) # histogram and qqplot for distribution of x

```

2\. Removal of outliers

```{r,fig.align='center', fig.dim=c(4,4)}
 
df<-data.frame(x,y) 
bag<-bagplot(df,show.whiskers = FALSE, ylim=c(0,20))
legend("topleft", c("bag","loop","outliers", "d.median"), 
           pch = c(15,15,16,8),col=c("blue","lightblue","red","red"),cex=0.7)

# Combine data points from "bag" and within the loop

dat<-rbind(bag$pxy.bag,bag$pxy.outer)

# Output is a matrix, we can pull out x and y variables for next stage 

x<-dat[,1] 
y<-dat[,2]
```

3\. Testing data for presence of boundary

This was already done using P and boundary existence was confirmed

### Fitting the boundary line to data

From the exploratory tests, they indicated that the data provides evidence of boundary existence, outliers have been identified and removed, and the variables *`x`* and *`y`* are normally distributed. We therefore, proceed to fit a boundary line model to the data set using the censored bivariate normal model. The function *`cbvn()`* fits the boundary line to the data.

We determine the argument values for the function *`cbvn()`* as before . Firstly, create a data-frame containing *`x`* and *`y`,* called `vals`.

```{r}
vals<-data.frame(x,y) #This is an input dataframe containing the variables
```

Secondly, the *`cbvn()`* requires initial starting values, *`theta`*, which are parameters of the boundary line (the censor) and the bivariate normal distribution. In this case, we shall fit a linear+Plateau model (lp) and hence the parameters are the plateau value and the intercept and slope of the linear component. With a scatter plot of *`y`* against *`x`* active in the plot window in `R`, run the function *`start.values("lp")`*, then click on the plot, the point you expect to be the lowest and the largest response of a linear model at the boundary. Note that starting values for other models can be determined by changing the model name from "lp" to the desired models (see *`?startValues`* help file)

```{r, eval=FALSE}
plot(x,y)

startValues("lp") 
```

Next, determine the parameters of the bivariate normal distribution which include the means of the *`x`* and *`y`* variables, standard deviation of the *`x`* and *`y`* variables, and the correlation of *`x`* and *`y`*.

```{r}
mean(x) 
mean(y) 
sd(x) 
sd(y) 
cor(x,y)

#The parameters of the boundary line and the data can be combined in a #vector theta in the order 
#theta<-c(intercept, slope, max response, mean(x), mean(y), sd(x), sd(y), cor(x,y))

theta<-c(-9,3, 13.5,7.5,9,0.68,2.3,0.12)
```

The standard deviation of the measurement error, *`sigh`* was already determine as 0.7. Now that all the arguments required for `cbvn()` function are set, the boundary line can be fitted

```{r, eval=TRUE,fig.align='center', fig.dim=c(4,4) }

model2<-cbvn(vals,theta,sigh=0.7,model = "lp", xlab=expression("pH"), ylab=expression("Yield /t ha"^{-1}), pch=16, col="grey")

model2
```

The boundary line is fitted and the parameters together with their corresponding standard error values are obtained. From our output, the `AIC` value of the BL model is lower than that of the bivariate normal model. Therefore, the BL model is appropriate. The parameters of the boundary line can now be used to predict the boundary yield given the value of pH using the function `predictBL()`. We need to predict the largest expected yield for each farm given the soil pH. We will replace any missing values with the mean value of pH.

```{r, eval=TRUE}

xpH<-soil$pH # let xpH be the P content in our dataset
xpH[which(is.na(xpH)==T)]<-mean(xpH,na.rm=T)
pH<-predictBL(model2,xpH)
```

## Determination of most limiting factor

The boundary line analysis has been used to determine the most limiting factor in many yield gap analysis studies. This is according to the von Liebig (1840) law of the minimum

$$
y=min(f_1(x_1),f_2(x_2),...,f_n(x_n))
$$

In the BLA package, the function `limfactor()` can be used to determine the most limiting factor at each point. For more information on the function, see

```{r, eval=FALSE}
?limfactor
```

For the two soil properties in our dataset, the most limiting factor for each data point is determined as

```{r, eval=TRUE}

Limiting_factor<-limfactor(P,pH) # This produces a list of length 2 containg a vector limiting factors at each pont and the maximum predicted response in the dataset.

Limiting_factors<-Limiting_factor[[1]] 
```

We can check the proportion of cases in which each of the soil properties was limiting

```{r, eval=TRUE,fig.align='center', fig.dim=c(4,4)}
Lim_factors<-Limiting_factors$Lim_factor

barplot(prop.table(table(Lim_factors))*100,
        ylab = "Percentage (%)",
        xlab = "Soil property",
        col = "grey",
        ylim=c(0,90))

axis(side = 1, at = seq(0, 4, by = 1), labels = FALSE, lwd = 1, col.ticks = "white")  
axis(side = 2, lwd = 1)  
```

As only two factors were evaluated, most of the yield gap cause was unidentified. This is because there many other factors affecting yield which we did not consider in this case. You can also plot the predicted yield by the most limiting factor against the actual yield.

```{r, eval=TRUE,fig.align='center', fig.dim=c(4,4)}

plot(Limiting_factors$Rs, soil$yield,
     xlab="Predicted yield (ton/ha)",
     ylab="Actual yield (ton/ha)", pch=16, col="grey")

abline(h=Limiting_factor[[2]], col="blue", lty=5, lwd=1)
lines(c(min(Limiting_factors$Rs),max(Limiting_factors$Rs)), 
c(min(Limiting_factors$Rs),max(Limiting_factors$Rs)), 
col="red", lwd=2)

legend("bottomleft",legend = c("Att yield", "1:1 line"),
       lty=c(5,1), col=c("blue", "red"), lwd=c(1, 2), cex = 0.8)
```

The horizontal red line represents the attainable yield for this particular setting. The 1:1 describes the situations in which the actual yield is equal to the predicted yield by the most limiting factor. The yield gap for each point is equivalent to the vertical distance between the point and the 1:1 line. The points above the 1:1 line are considered to have zero yield gap and their position above the line is only due to measurement error.

## Concluding remarks

The censored bivariate normal model has been used to carryout a yield gap analysis. The advantage of using this method over the other boundary line fitting methods is that it is based on statistical principals and it removes the subjectivity associated with selection of bin sizes and the quantile to consider as boundary as it is done in the binning methodology and quantile regression. Furthermore, the uncertainty around the boundary line can also be determined.

# References

1\. Milne, A. E., Wheeler, H. C., & Lark, R. M. (2006). On testing biological data for the presence of a boundary. Annals of Applied Biology, 149 , 213-222. doi:<https://doi.org/10.1111/j.1744-7348.2006.00085.x>

2\. Lark, R. M., Gillingham, V., Langton, D., & Marchant, B. P. (2020). Boundary line models for soil nutrient concentrations and wheat yield in national-scale datasets.European Journal of Soil Science, 71 , 334-351. doi: <https://doi.org/10.1111/>[ejss.12891](https://doi.org/10.1111/ejss.12891)

3\. Lark, R. M., & Milne, A. E. (2016). Boundary line analysis of the effect of water filled pore space on nitrous oxide emission from cores of arable soil. European Journal of Soil Science, 67 , 148-159. doi: <https://doi.org/10.1111/ejss.12318>
